RL-Musician: A Tool for Music Composition with Deep Reinforcement Learning

### 2019-10-06 docs/paper/paper.tex
docs: Continue writing paper draft

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography

@@

Brief outline of the approach is as follows. There is an environment with 2D table representing piano roll which means that rows of the table correspond to consecutive notes and columns of the table correspond to time steps. An agent observes current time step and some previous time steps. An action of the agent can be either playing a note at the current time step or moving to the next time step. When the last time step of the piano roll is reached, episode is ended and the agent receives reward based on evaluation of created composition. Trainable parameters of the agent are weights of a neural network used as a so called actor model, i.e., a model that maps observation to probabilities of actions. These parameters are trained with cross-entropy method (alternatively, they can be trained with genetic algorithms, ant colony algorithms, or evolutionary strategies \cite{salimans2017evolution}).

@@

\section{Methodology}
\label{sec:methodology}
% dataset -> rules; backpropagation -> cross-entropy; learning rate -> weights std.

>>>>

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{algorithm}      % environment for algorithms
\usepackage{algorithmic}    % pseudocode

@@

Brief outline of the approach is as follows. There is an environment with 2D table representing piano roll which means that rows of the table correspond to consecutive notes and columns of the table correspond to time steps. An agent observes current time step and some previous time steps. An action of the agent can be either playing a note at the current time step or moving to the next time step. When the last time step of the piano roll is reached, episode is ended and the agent receives reward based on evaluation of created composition. Trainable parameters of the agent are weights of a neural network used as a so called actor model, i.e., a model that maps observation to probabilities of actions. These parameters are trained with cross-entropy method \cite{rubinstein1997optimization} (alternatively, they can be trained with genetic algorithms, ant colony algorithms, or evolutionary strategies \cite{salimans2017evolution}).

@@

\begin{algorithm}
	\caption{Cross-entropy method for optimization} \label{alg:crossentropy}
	\textbf{Input:} $X$ -- set of elements, $f: X \to \mathbb{R}$ -- target function, $u(\cdot, w)$ -- probabilistic distribution over $X$ parametrized by vector $w$. \\
	\textbf{Output:} $\hat{w}$ -- approximate solution to the problem $\max_w \mathbb{E}_{x \sim u(\cdot, w)} f(x)$. \\
	\textbf{Hyperparameters:} $w_0$ -- initial value of $w$; $N$ -- number of iterations, $n$ -- number of vectors to draw at each iteration, $\sigma$ -- standard deviation for vectors generation, $m$ -- number of trials for each vector; $\rho$ -- fraction of best vectors to use for update; $\alpha$ -- smoothing coefficient of updates.
	\begin{algorithmic}[1]
		\FORALL{$i \in \{1, \dots, N\}$}
		    \FORALL{$j \in \{1, \dots, n\}$}
		        \STATE{draw $w_{ij} \sim \mathcal{N}(\cdot \vert w_{i-1}, \sigma)$}
		        \STATE{$r_j \gets \sum_{k = 1}^{m} f(x_k)$ where $x_k \sim u(\cdot, w_{ij})$}
		    \ENDFOR
		    \STATE{$r_{\mathrm{threshold}} \gets$ $[\rho n]$-th highest value of $\{r_j: j \in \{1, \dots, n\}\}$}
		    \STATE{$J \gets \{j: r_j \ge r_{\mathrm{threshold}}\}$}
		    \STATE{$w_i \gets \alpha w_{i-1} + (1 - \alpha)(\sum_{j \in J} w_{ij}) / [\rho n]$}
		\ENDFOR
	    \STATE{$\hat{w} \gets w_N$}
	\end{algorithmic}
\end{algorithm}

Algorithm \ref{alg:crossentropy} defines a variant of cross-entropy method for optimization. Sometimes, hyperparameter $m$ is omitted and intermediate results are not aggregated over multiple trials. In case of $u(\cdot, w)$ that acts like a deterministic function of $w$, $m$ is redundant, but, in general case, terminal result can be improved by setting $m > 1$.

\section{Methodology}
\label{sec:methodology}

\subsection{Setup}
\label{subsec:setup}

For this study, piano roll is chosen as music representation format, but there are some subtleties.

Here, each row of a table does not correspond to a particular note or pitch. Relative axis is used instead of absolute axis. All that is important is that each row is one semitone higher than the row right below it. Thus, a musical piece is given only up to transposition and user-defined value for a pivot note is required to define absolute pitches.

\subsection{Evaluational Rules}
\label{subsec:setup}
